Foundations of a live data exploration environment

### 2020-02-13 wip-live-programming/paper.tex
Final version edits

\documentclass[english,submission]{programming}
\usepackage[backend=biber]{biblatex}
\addbibresource{paper.bib}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumitem}

\newcounter{thc}
\newcounter{dfc}

\theoremstyle{plain}
\newtheorem{lem}[thc]{Lemma}
\newtheorem{theorem}[thc]{Theorem}

@@

\begin{document}
\title{A live coding environment for a data exploration language}
%\titlerunning{Learning to live with errors}

\author[a]{Tomas Petricek}
\authorinfo{
\noindent
Tomas is a lecturer at University of Kent. He is working on making programming and data science
easier and more accessible. His research interests span a range of areas including theory of
programming languages, history and philosophy of programming, as well as open data and data
journalism.
%
Previously, he contributed to functional programming and the development of the F\# language and
type providers at Microsoft Research and obtained PhD from University of Cambridge for his work
on coeffects, a theory of context-aware programming languages. Along the way, he became interested
in understanding programming through the perspective of history and philosophy of science and wrote
papers about the evolution of programming concepts such as types and errors.
%
Contact him at \email{tomas@tomasp.net}.}
\affiliation[a]{University of Kent, UK}
\keywords{Data exploration, live coding, data journalism, instant previews}
\paperdetails{
  perspective=sciencetheoretical,
  area={Programming environments},
}

% https://dl.acm.org/ccs/ccs.cfm and generate your Classification
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002951.10003227.10003351</concept_id>
<concept_desc>Information systems~Data mining</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041</concept_id>
<concept_desc>Software and its engineering~Compilers</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003121.10003129</concept_id>
<concept_desc>Human-centered computing~Interactive systems and tools</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{Human-centered computing~Interactive systems and tools}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Software and its engineering~Compilers}
\maketitle

\vspace{1em}
\noindent
{\scriptsize
~\\
\textbf{Note to reviewers:} This paper has been previously submitted to a different
venue. The revisions made in response to past reviews are summarized in Appendix~\ref{sec:revisions}
(to be removed for publication).}
\vspace{2em}

@@

An increasing number of programming environments aim to provide a live development experience
for standard programming languages, but doing this is not easy. Fully recomputing the whole program
after every keystroke is inefficient and calculating how a change in the source code changes the
result is extremely hard when the text editor allows arbitrary changes.
Consider the following snippet that gets the release years of the $10$ most expensive movies from a
data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
\end{array}
\end{equation*}
%
A live coding environment computes and displays the list of years. Suppose that the programmer then
makes $\num{10}$ a variable and changes the date format:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\ident{count}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking the first 10 elements) and only evaluate the
\ident{map} operation to differently format the release dates of top 10 movies. Our environment
does this for a simple data exploration language in an unrestricted text editor. We discuss related
work in Section~\ref{sec:future}, but we briefly review the most important directions here, in
order to situate our contributions.

\subsection{Related work}
Simple data exploration performed, for example, by journalists \cite{ddj} is done either
programmatically or using spreadsheets. The latter is easy but error-prone while the former
requires expert programming skills. We aim to bring liveness of spreadsheets to programmatic
data exploration. This requires extending recomputation as done in spreadsheets \cite{spreadsheet}
to code written in an ordinary text editor.

Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some
can even export the workflow as a script \cite{wrangler}. Data analysts who prefer working with
code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
Those are text-based, but have a limited model of recomputation. Users structure code in
cells and manually reevaluate cells. Many notebook systems do not track dependencies
between cells which can lead to well-documented inconsistencies \cite{dataflow,noworkflow,wrattler}.

Existing text-based editors that provide instant feedback,
such as Lighttable \cite{lighttable} and Chrome DevTools, often work only in certain
situations or require full recomputation. A more principled approach can be used by
systems based on structured editing \cite{livenut,lamdu} where code is only
modified via known operations with known effect on the computation graph
(e.g.~``extract variable'' has no effect on the result; ``change constant value''
forces recomputation of subsequent code). We aim to cater for the many users who
prefer to edit programs as free-form text.
%
The technical aspects of our implementation are related to incremental computation \cite{selfadjusting,incremental}.
This focuses on cases where \emph{data changes} but program stays the same, while our focus is on cases where
\emph{program changes} but data stays the same.

\subsection{Contributions}
We present the design and implementation of a live coding environment for a simple data exploration
language that provides correct and efficient instant feedback, yet is integrated into an
ordinary text editor. Our key contributions are:

@@

\item A live coding environment does not operate in batch mode and so we cannot follow classic
  compiler literature. We capture the essence of such new perspective (Section~\ref{sec:formal})
  and use it to build our \emph{live preview} mechanism (Section~\ref{sec:previews}) that evaluates
  code instantly during editing.

\item We prove that our live preview mechanism is correct (Section~\ref{sec:evaluation-correctness}) and that
  it reuses previously evaluated values when possible (Section~\ref{sec:evaluation-reuse}). We
  illustrate the practicality of the mechanism using an empirical evaluation
  (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:evaluation-case}).
\end{itemize}

@@

\begin{figure}
\vspace{-1em}
\hspace{1em}\includegraphics[scale=0.52]{figures/notebook.png}
\vspace{-0.5em}
\caption{Financial Times analysis that joins UN trade database with ISO country codes.}
\label{fig:ft-uncomtrade}
\vspace{-0.5em}
\end{figure}

@@

Consider the Financial Times analysis of plastic waste \cite{ftnotebooks,ftarticle}. It joins
datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an excerpt from one notebook of the data analysis. The code has a number of important properties:
%
\begin{itemize}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any custom functions. Code is parameterized by having a global variable
  \ident{material} set to \str{plastics} and keeping other possible values in a comment.
  This lets the analyst run and check results of intermediate steps.

\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides operations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Such standard libraries
  are external to the data analysis and are often implemented in another language like C++.

\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\item There are many corner cases, such as the fact that \ident{keep\_default\_na}
  needs to be set to handle Namibia correctly. These are discovered interactively by
  running the code and examining the output, so providing an instant feedback is essential.
\end{itemize}

\noindent
Many Jupyter notebooks are more complex than the above example and might use helper
functions or object-oriented code. However, simple data analyses such as the one discussed
here are frequent enough and pose interesting problems for programming tools.
This paper aims to bring such analyses to the attention of programming research community
by capturing their essential properties as a formal calculus.

@@

\begin{figure}
  \vspace{-1.5em}
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The analysis counts the number of distinct events per
      athlete. After typing `.' the editor offers further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{figures/gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize
      Our live coding environment for The Gamma. The table is updated on-the-fly a
      and shows the result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{figures/gamma2.png}}
  \caption{\small Previous work on The Gamma (left) and our new editor with live preview (right).}
  \label{fig:gamma-intro}
  \vspace{-0.5em}
\end{figure}

@@

Simple data exploration as discussed in the previous section has been the motivation for
a scripting language The Gamma \cite{gamma}. Scripts in The Gamma are sequences of commands
that either define a variable or produce an output. It does not support top-level
functions and lambda functions can be used only as method arguments.

Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
such as JavaScript. The Gamma uses type providers \cite{providers-fsharp}
for accessing and exploring external data sources.
Type providers generate object types with members and The Gamma makes using those convenient by
providing auto-complete when the user types the dot symbol (`.') to access a member.
The combination of type providers and auto-complete makes it
possible to solve a large number of data exploration tasks through the very simple interaction of
selecting operations from a list.

@@

\subsection{Live coding for data exploration}

The implementation that accompanies this paper builds a live coding environment for The Gamma.
The implementation is discussed in Section~\ref{sec:evaluation-case} and it replaces
the original text editor with just auto-complete with a live coding environment
that provides \emph{live previews}. We choose to extend The Gamma over using a subset of Python
as the simplicity of TheGamma lets us focus on the principles of live coding systems.

An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma programs consist of lists of commands which are either expressions or let bindings. Our
editor displays a live preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

@@

The \emph{data exploration calculus} is a small formal language for data exploration. The calculus
is not, in itself, Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
define the calculus in this section and then use it to formalise our live preview
mechanism in Section~\ref{sec:formal}.
%
The live preview mechanism does not rely on types and so we postpone the discussion of static
typing to Section~\ref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
previews.

@@

\begin{figure}
\vspace{-0.5em}
\raggedright
\hspace{0.5em}{\small\sffamily Programs, commands, terms, expressions and values}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{lcl}
p &\narrow{::=}& c_1; \ldots; c_n\\
c &\narrow{::=}& \kvd{let}~x = t \lsep t\\
\end{array}
\qquad
\begin{array}{lcl}
t &\narrow{::=}& o \lsep x \\
  &\narrow{|}& t.m(e, \ldots, e)
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{::=}& t \lsep \lambda x\rightarrow e\\
v &\narrow{::=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

@@

\vspace{0.5em}
\hspace{0.5em}{\small\sffamily Let elimination and member reduction}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}
o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
\qquad  o_1;\; \ldots;\; o_{k};\; o; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\rname{let}}\\[1.2em]
{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
\quad\text{\rname{external}}
\end{array}
\end{equation*}

\vspace{-0.5em}
\caption{Syntax, contexts and reduction rules of the data exploration calculus}
\label{fig:dec-calculus}
\vspace{-0.5em}
\end{figure}

@@

The data exploration calculus is a call-by-value language. We model evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines the reduction behaviour for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Here, the operation $m$ is invoked on an object and takes values (objects or function values) as arguments.
The reduction always results in an object.
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evaluation of a command and $C_p$ defines left-to-right evaluation of a program.
The rule \rname{external} calls a method provided by an external library in a
call-by-value fashion and \rname{let} substitutes a value of an evaluated variable in
all subsequent commands and leaves the result in the list of commands.
Note that our semantics does not define how $\lambda$ functions are reduced. This is done by
external libraries, which will typically supply functions with arguments using standard $\beta$-reduction.
The behaviour is subject to constraints discussed next.

\subsection{Properties}
The data exploration calculus has a number of desirable properties. Some of those require that the
relation $\rightsquigarrow_\epsilon$, which defines evaluation for external libraries, satisfies a
number of conditions. We discuss \emph{normalization} and \emph{let elimination} in this section.
Those two are particularly important as they will allow us to prove correctness of our method of
evaluating live previews in Section~\ref{sec:evaluation-correctness}.

@@

\noindent
We say that two expressions
$e$ and $e'$ are \emph{observationally equivalent} if, for any context $C$, the expressions $C[e]$ and
$C[e']$ reduce to the same value. Lambda functions $\lambda x\!\rightarrow\!2$ and
$\lambda x\!\rightarrow\!1\!+\!1$ are not equal, but they are observationally equivalent.
We require that external libraries satisfy two conditions. First, when a method is called with
observationally equivalent values as arguments, it should return the same value. Second, the
evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, n$ and $v_i$. The external
library will typically satisfy this by defining an error object and reducing all invalid calls
to the error object \cite{gowrong}.

@@

\noindent
Compositionality is essential for proving the correctness of our live preview mechanism.
Totality allows us to prove normalization, i.e.~all programs reduce to
a value -- although the resulting value may be an error value provided by the external library.

@@

\noindent
Although the reduction rules \rname{let} and \rname{external} of the data exploration calculus
define an evaluation in a call-by-value order, eliminating let bindings in a call-by-name way
using the $\rightsquigarrow_\ident{let}$ reduction does not affect the result. This
simplifies our later proof of live preview correctness in Section~\ref{sec:evaluation-correctness}.

@@

\section{Formalising a live coding environment}
\label{sec:formal}

A naive way of providing live previews during code editing would be to re-evaluate the code
after each change. This would be wasteful -- when writing code to explore data, most changes are
additive. To update a preview, we only need to evaluate newly added code. We describe an
efficient mechanism in this section.

@@

\subsection{Maintaining dependency graph}
\label{sec:formal-deps}

@@

\begin{figure}
\centering
\subcaptionbox{
  \raggedright\small
  Graph constructed from initial expression:
  $\kvd{let}~x = \num{15}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
  \label{fig:dep-graph-a}
}{
  $\xymatrix{
  \bnd{val}(\num{10}) & \bnd{val}(\num{15})\\
  \bnd{mem}(\ident{skip},s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
  \bnd{mem}(\ident{take},s_1)\ar[l]_{\blbl{arg}(0)} \ar[u]_{\blbl{arg}(1)}\\
  \bnd{var}(\ident{data})
  }\qquad$
}
\hspace{1em}
\subcaptionbox{
  \raggedright\small
  Updated graph after changing $x$ to $10$:
  $\kvd{let}~x = \num{10}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
  \label{fig:dep-graph-b}
}{
  $\xymatrix{
  \bnd{val}(\num{10})&\\
  \bnd{mem}(\ident{skip}, s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
  \bnd{mem}(\ident{take}, s_2)\ar[l]_{\blbl{arg}(0)} \ar@/_/[lu]_{\blbl{arg}(1)}\\
  \bnd{var}(\ident{data})
}\qquad$
}
\caption{Dependency graphs formed by two steps of the live programming process. }
\label{fig:dep-graph}
\end{figure}

@@

\section{Computing live previews}
\label{sec:previews}

@@

\section{Evaluating live coding environment}
\label{sec:evaluation}

@@

To avoid clutter when visualizing the performance, we update the preview after complete tokens are
added rather than after individual keystrokes.
Figure~\ref{fig:image-edits} shows the sequence of edits that we use to measure
the delays in updating a live preview. We first enter an expression to load, greyscale and blur
an image (1) then introduce let binding (2) and add more operations (3). Finally, we extract
one of the parameters into a variable (4). Most of the operations are simply adding code, but
there are two cases where we modify existing code and change value of a parameter for \ident{blur}
and \ident{combine} immediately after (1) and (3), respectively.

>>>>

\documentclass[english,crc]{programming}
\usepackage[backend=biber]{biblatex}
\addbibresource{paper.bib}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumitem}

\newcounter{thc}
\newcounter{dfc}
\theoremstyle{plain}
\newtheorem{lem}[thc]{Lemma}
\newtheorem{theorem}[thc]{Theorem}

@@

\begin{document}
\title{Foundations of a live data exploration environment}
%\titlerunning{Learning to live with errors}
\author[a,b]{Tomas Petricek}
\authorinfo{
\noindent
Tomas is a lecturer at University of Kent. He is working on making programming and data science
easier and more accessible. His research interests span a range of areas including theory of
programming languages, tools for data science as well as history and philosophy of programming.
%
Previously, he contributed to functional programming and the development of the F\# language and
type providers at Microsoft Research and obtained PhD from University of Cambridge for his work
on coeffects, a theory of context-aware programming languages. Along the way, he became interested
in understanding programming through the perspective of history and philosophy of science and wrote
papers about the evolution of programming concepts such as types and errors.
%
Contact him at \email{tomas@tomasp.net}.}
\affiliation[a]{University of Kent, UK}
\affiliation[b]{and The Alan Turing Institute, UK}
\keywords{Data exploration, live programming, data journalism, instant previews}
\paperdetails{
  perspective=sciencetheoretical,
  area={Programming environments},
}
% https://dl.acm.org/ccs/ccs.cfm and generate your Classification
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002951.10003227.10003351</concept_id>
<concept_desc>Information systems~Data mining</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041</concept_id>
<concept_desc>Software and its engineering~Compilers</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003121.10003129</concept_id>
<concept_desc>Human-centered computing~Interactive systems and tools</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Human-centered computing~Interactive systems and tools}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Software and its engineering~Compilers}
\maketitle

@@

An increasing number of programming environments aim to provide a live development experience
for standard programming languages, but doing this is not easy. Fully recomputing the whole program
after every keystroke is inefficient and calculating how a change in the source code changes the
result is extremely hard when the text editor allows arbitrary changes.
Consider the following snippet that gets the release years of the $10$ most expensive movies from a
data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} =\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))\\
\end{array}
\end{equation*}
%
A live programming environment computes and displays the list of years. Suppose that the programmer then
makes $\num{10}$ a variable and changes the date format:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\\
\qquad .\ident{take}(\ident{count}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))\\
\end{array}
\end{equation*}
%
Ideally, the live programming environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking the first 10 elements) and only evaluate the
\ident{map} operation to differently format the release dates of top 10 movies. Our environment
does this for a simple data exploration language in an unrestricted text editor. We discuss related
work in Section~\ref{sec:future}, but we briefly review the most important directions here, in
order to situate our contributions.

\subsection*{Contributions}
We present the design and implementation of a live programming environment for a simple data exploration
language that provides correct and efficient instant feedback, yet is integrated into an
ordinary text editor. Our key contributions are:

@@

\item A live programming environment does not operate in batch mode and so we cannot follow classic
  compiler literature. We capture the essence of such new perspective (Section~\ref{sec:formal})
  and use it to build our \emph{instant preview} mechanism (Section~\ref{sec:previews}) that evaluates
  code instantly during editing.

\item We prove that our instant preview mechanism is correct (Section~\ref{sec:evaluation-correctness}) and that
  it reuses previously evaluated values when possible (Section~\ref{sec:evaluation-reuse}). We
  illustrate the practicality of the mechanism using an empirical evaluation
  (Section~\ref{sec:evaluation-empirical}) and a case study (Section~\ref{sec:evaluation-case}).
\end{itemize}

@@

\begin{figure}[b]
\hspace{1em}\includegraphics[scale=0.52]{figures/notebook.png}
\caption{Financial Times analysis that joins UN trade database with ISO country codes.}
\label{fig:ft-uncomtrade}
\end{figure}

@@

Consider the Financial Times analysis of plastic waste \cite{ftnotebooks,ftarticle}. It joins
datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an excerpt from one notebook of the data analysis. The code has a number of important properties:
%
\begin{itemize}
\vspace{0.5em}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any custom functions. Code is parameterized by having a global variable
  \ident{material} set to \str{plastics} and keeping other possible values in a comment.
  This lets the analyst run and check results of intermediate steps.

\vspace{0.5em}
\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides operations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Such standard libraries
  are external to the data analysis and are often implemented in another language like C++.

\vspace{0.5em}
\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\vspace{0.5em}
\item There are many corner cases, such as the fact that \ident{keep\_default\_na}
  needs to be set to handle Namibia correctly. These are discovered interactively by
  running the code and examining the output, so providing an instant feedback is essential.
\end{itemize}

\vspace{0.5em}
\noindent
Many Jupyter notebooks are more complex than the above example and might use helper
functions or object-oriented code. However, simple data analyses such as the one discussed
here are frequent enough and pose interesting problems for programming tools.
This paper aims to bring such analyses to the attention of programming research community
by capturing their essential properties as a formal calculus.

@@

\begin{figure}
\centering
\subcaptionbox{
  \raggedright\footnotesize The analysis counts the number of distinct events per
    athlete. After typing `.' the editor offers further aggregation operations.
  \label{fig:gamma-dot}
}{\includegraphics[scale=0.26]{figures/gamma1.png}}
\subcaptionbox{
  \raggedright\footnotesize
    Our live programming environment for The Gamma. The table is updated on-the-fly a
    and shows the result at the current cursor position.
  \label{fig:gamma-preview}
}{\includegraphics[scale=0.26]{figures/gamma2.png}}
\vspace{0.5em}
\caption{\small Previous work on The Gamma with auto-complete based on type information (left)
  and our new editor with instant preview (right).}
\label{fig:gamma-intro}
\end{figure}

@@

Simple data exploration has been the motivation for
a scripting language The Gamma \cite{gamma}. Scripts in The Gamma are sequences of commands
that either define a variable or produce an output. It does not support top-level
functions and lambda functions can be used only as method arguments.
Given the limited expressiveness of The Gamma, libraries are implemented in other languages,
such as JavaScript. The Gamma uses type providers \cite{providers-fsharp}
for accessing external data sources.
Type providers generate object types with members and The Gamma offers those in an
auto-complete list when the user types dot (`.') to access a member.
The combination of type providers and auto-complete makes it
possible to solve a large number of data exploration tasks through the very simple interaction of
selecting operations from a list.

@@

\subsection{Live programming for data exploration}

The implementation that accompanies this paper builds a live programming environment for The Gamma.
It is discussed in Section~\ref{sec:evaluation-case} and it replaces
the original text editor with just auto-complete with a live programming environment
that provides \emph{instant previews}.

An example of a instant preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma programs consist of lists of commands which are either expressions or let bindings. Our
editor displays a instant preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

@@

The \emph{data exploration calculus} is a small formal language for data exploration. The calculus
is not, in itself, Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
define the calculus in this section and then use it to formalise our instant preview
mechanism in Section~\ref{sec:formal}.
%
The instant preview mechanism does not rely on types and so we postpone the discussion of static
typing to Appendix~\ref{sec:extra-types}. Interestingly, it reuses the mechanism used for live
previews.

@@

\begin{figure}
\raggedright
\hspace{0.5em}{\small\sffamily Programs, commands, terms, expressions and values}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{lcl}
p &\narrow{::=}& c_1; \ldots; c_n\\
c &\narrow{::=}& \kvd{let}~x = t \lsep t\\
\end{array}
\qquad
\begin{array}{lcl}
t &\narrow{::=}& o \lsep x \\
  &\narrow{|}& t.m(e, \ldots, e)
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{::=}& t \lsep \lambda x\rightarrow e\\
v &\narrow{::=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

@@

\vspace{0.5em}
\hspace{0.5em}{\small\sffamily Let elimination and member reduction}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}
o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
\qquad  o_1;\; \ldots;\; o_{k};\; o; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\rname{let}}\\[1.2em]
\inference
  {o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
\quad\text{\rname{external}}
\end{array}
\end{equation*}

\caption{Syntax, contexts and reduction rules of the data exploration calculus}
\label{fig:dec-calculus}
\end{figure}

@@

The data exploration calculus is a call-by-value language. We model evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines the reduction behaviour for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Here, the operation $m$ is invoked on an object and takes values (objects or function values) as arguments.
The reduction always results in an object.
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evaluation of a command and $C_p$ defines left-to-right evaluation of a program.
The rule \rname{external} calls a method provided by an external library in a
call-by-value fashion; \rname{let} substitutes a value of an evaluated variable in
all subsequent commands and leaves the result in the list of commands.
Note that our semantics does not define how $\lambda$ applications are reduced. This is done by
external libraries, which will typically supply functions with arguments using standard $\beta$-reduction.
The behaviour is subject to constraints discussed next.

\subsection{Example external library}
\label{sec:calculus-example}

The data exploration calculus is not limited to the data exploration domain.
It can be used with external libraries for a wide range of other simple programming tasks, such as image
manipulation, as done in Section~\ref{sec:evaluation-empirical}. However, we choose a name that reflects the
domain that motivated this paper. To illustrate how a definition of an external library looks,
consider the following simple data manipulation script:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~l = \ident{list}.\ident{range}(\num{0}, \num{10})\\
l.\ident{map}(\lambda x \rightarrow \ident{math}.\ident{mul}(x, \num{10}))
\end{array}
\end{equation*}
%
An external library provides the \ident{list} and \ident{math} values with members
\ident{range}, \ident{map} and \ident{mul}. The objects of the external library consist of
numbers $n$, lists of objects $[o_1, \ldots, o_k]$ and failed computations $\bot$ \cite{gowrong}.
Next, the external library needs to define the $\rightsquigarrow_\epsilon$ relation that defines
the evaluation of member accesses. The following shows the rules for members of lists, assuming
the only supported member is \ident{map}:
%
\begin{equation*}
\begin{array}{l}
\\[-0.75em]
\inference
  {e[x \leftarrow n_i] \rightsquigarrow o_i \quad (\textit{for all $i\in 1\ldots k$})}
  {[ n_1, \ldots, n_k ].\ident{map}(\lambda x\rightarrow e) \rightsquigarrow_\epsilon [ o_1, \ldots, o_k ]}
\qquad
\inference
  {(\textit{otherwise})}
  {[ n_1, \ldots, n_k ].m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon \bot}
\\[-0.75em]
~
\end{array}
\end{equation*}
%
When evaluating \ident{map}, we apply the provided function to all elements of the list using
standard $\beta$-reduction, defined recursively using $\rightsquigarrow$, and return a list with
resulting objects. The $\rightsquigarrow_\epsilon$ relation is defined on all member accesses,
but non-existent members reduce to the failed computation $\bot$.

\subsection{Properties}
The data exploration calculus has a number of desirable properties. Some of those require that the
relation $\rightsquigarrow_\epsilon$, which defines evaluation for external libraries, satisfies a
number of conditions. We discuss \emph{normalization} and \emph{let elimination} in this section.
Those two are particularly important as they will allow us to prove correctness of our method of
evaluating instant previews in Section~\ref{sec:evaluation-correctness}.

@@

\noindent
We say that two expressions
$e$ and $e'$ are \emph{observationally equivalent} if, for any context $C$, the expressions $C[e]$ and
$C[e']$ reduce to the same value. Lambda functions $\lambda x\!\rightarrow\!2$ and
$\lambda x\!\rightarrow\!1\!+\!1$ are not equal, but they are observationally equivalent.
We require that external libraries satisfy two conditions. First, when a method is called with
observationally equivalent values as arguments, it should return the same value. Second, the
evaluation of $o.m(v_1,\ldots,v_n)$ should be defined for all $o, n$ and $v_i$.
The definition in Section~\ref{sec:calculus-example} satisfies those by using standard
$\beta$-reduction for lambda functions and by reducing all invalid calls to the $\bot$ object.

@@

\noindent
Compositionality is essential for proving the correctness of our instant preview mechanism and
implies determinism of external libraries.
Totality allows us to prove normalization, i.e.~all programs reduce to
a value -- although the resulting value may be an error value provided by the external library.

@@

\noindent
Although the reduction rules \rname{let} and \rname{external} of the data exploration calculus
define an evaluation in a call-by-value order, eliminating let bindings in a call-by-name way
using the $\rightsquigarrow_\ident{let}$ reduction does not affect the result. This
simplifies our later proof of instant preview correctness in Section~\ref{sec:evaluation-correctness}.

@@

\section{Formalising a live programming environment}
\label{sec:formal}

A naive way of providing instant previews during code editing would be to re-evaluate the code
after each change. This would be wasteful -- when writing code to explore data, most changes are
additive. To update a preview, we only need to evaluate newly added code. We describe an
efficient mechanism in this section.

@@

\subsection{Maintaining the dependency graph}
\label{sec:formal-deps}

@@

@@

\section{Computing instant previews}
\label{sec:previews}

@@

\section{Evaluating live programming environment}
\label{sec:evaluation}

@@

To avoid distractions when visualizing the performance, we update the preview after complete tokens are
added rather than after individual keystrokes.
Figure~\ref{fig:image-edits} shows the sequence of edits that we use to measure
the delays in updating a instant preview. We first enter an expression to load, greyscale and blur
an image (1) then introduce let binding (2) and add more operations (3). Finally, we extract
one of the parameters into a variable (4). Most of the operations are simply adding code, but
there are two cases where we modify existing code and change value of a parameter for \ident{blur}
and \ident{combine} immediately after (1) and (3), respectively.
